---
title: "p8105_hw2_xf2302"
output: github_document
date: "2025-09-30"
---
# Problem 1
```{r}
library(tidyverse)
```
# Clean pols-month.csv dataset
```{r}
pols <- 
  read_csv("data/pols-month.csv") |> 
  separate(mon, into = c("year", "month", "day"), sep = "-") |> 
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    month = month.name[month],
    president = if_else(prez_gop == 1, "gop", "dem")
  ) |> 
  select(-day, -prez_gop, -prez_dem)
```
# 2. Clean snp.csv dataset
```{r}
snp <- 
  read_csv("data/snp.csv") |> 
  separate(date, into = c("month", "day", "year"), sep = "/") |> 
  mutate(
    month = as.integer(month),
    year = as.integer(year),
    year = if_else(year < 100,            
                   if_else(year >= 50, 1900 + year, 2000 + year),
                   year),
    month = month.name[month]
  ) |> 
  select(year, month, close) |> 
  arrange(year, month)
```
# 3. Clean unemployment.csv
```{r}
unemployment <- 
  read_csv("data/unemployment.csv") |> 
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemployment"
  ) |> 
  mutate(
    year = as.integer(Year),
    month = month.name[match(month, month.abb)]
  ) |> 
  select(year, month, unemployment)
```
# 4. Merge datasets and inspect merged dataset
```{r}
merged_data <- 
  pols |> 
  left_join(snp, by = c("year", "month")) |> 
  left_join(unemployment, by = c("year", "month"))
dim(merged_data)        
range(merged_data$year) 
head(merged_data)       
```
The **FiveThirtyEight datasets** combine information on U.S. politics, the stock market, and unemployment. 
The `pols-month.csv` dataset contains monthly counts of national politicians (governors, senators, and representatives) 
by political party, from `r min(pols$year)` to `r max(pols$year)`. 

The `snp.csv` dataset records monthly closing values of the S&P 500 index, ranging from `r min(snp$year)` to `r max(snp$year)`. 
The `unemployment.csv` dataset provides national unemployment rates, reported monthly from `r min(unemployment$year)` to `r max(unemployment$year)`.  

After cleaning and merging these datasets by year and month, the resulting dataset has **`r nrow(merged_data)` rows** 
and **`r ncol(merged_data)` variables**, covering the period **`r min(merged_data$year)`–`r max(merged_data$year)`**. 
Key variables include the counts of politicians by party, the sitting president’s party, monthly unemployment rates, 
and the S&P 500 closing values. Note that missing values occur where data are not available 
(e.g., stock market values before `r min(snp$year)`, unemployment data in `r min(unemployment$year)-1`).

