---
title: "p8105_hw2_xf2302"
output: github_document
date: "2025-09-30"
---
# Problem 1
```{r}
library(tidyverse)
```
# Clean pols-month.csv dataset
```{r}
pols <- 
  read_csv("data/pols-month.csv") |> 
  separate(mon, into = c("year", "month", "day"), sep = "-") |> 
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    month = month.name[month],
    president = if_else(prez_gop == 1, "gop", "dem")
  ) |> 
  select(-day, -prez_gop, -prez_dem)
```
# 2. Clean snp.csv dataset
```{r}
snp <- 
  read_csv("data/snp.csv") |> 
  separate(date, into = c("month", "day", "year"), sep = "/") |> 
  mutate(
    month = as.integer(month),
    year = as.integer(year),
    year = if_else(year < 100,            
                   if_else(year >= 50, 1900 + year, 2000 + year),
                   year),
    month = month.name[month]
  ) |> 
  select(year, month, close) |> 
  arrange(year, month)
```
# 3. Clean unemployment.csv
```{r}
unemployment <- 
  read_csv("data/unemployment.csv") |> 
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemployment"
  ) |> 
  mutate(
    year = as.integer(Year),
    month = month.name[match(month, month.abb)]
  ) |> 
  select(year, month, unemployment)
```
# 4. Merge datasets and inspect merged dataset
```{r}
merged_data <- 
  pols |> 
  left_join(snp, by = c("year", "month")) |> 
  left_join(unemployment, by = c("year", "month"))
dim(merged_data)        
range(merged_data$year) 
head(merged_data)       
```
The **FiveThirtyEight datasets** combine information on U.S. politics, the stock market, and unemployment. 
The `pols-month.csv` dataset contains monthly counts of national politicians (governors, senators, and representatives) 
by political party, from `r min(pols$year)` to `r max(pols$year)`. 

The `snp.csv` dataset records monthly closing values of the S&P 500 index, ranging from `r min(snp$year)` to `r max(snp$year)`. 
The `unemployment.csv` dataset provides national unemployment rates, reported monthly from `r min(unemployment$year)` to `r max(unemployment$year)`.  

After cleaning and merging these datasets by year and month, the resulting dataset has **`r nrow(merged_data)` rows** 
and **`r ncol(merged_data)` variables**, covering the period **`r min(merged_data$year)`–`r max(merged_data$year)`**. 
Key variables include the counts of politicians by party, the sitting president’s party, monthly unemployment rates, 
and the S&P 500 closing values. Note that missing values occur where data are not available 
(e.g., stock market values before `r min(snp$year)`, unemployment data in `r min(unemployment$year)-1`).

# Problem 2
```{r}
library(readxl)
library(dplyr)
library(janitor)
```

# Define a Cleaning Function
```{r}
clean_trash <- function(sheet_name, wheel_name) {
  df <- read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx",
    sheet = sheet_name,
    skip = 1
  ) |>
    clean_names() |> 
    filter(!is.na(dumpster)) |>          
    select(-starts_with("...")) |> 
    mutate(
      year = as.integer(year)   
    )
  
  if ("sports_balls" %in% colnames(df)) {
    df <- df |>
      mutate(sports_balls = as.integer(round(sports_balls, 0)))
  } else {
    df <- df |>
      mutate(sports_balls = NA_integer_)
  }
  df |>
    mutate(wheel = wheel_name)
}
```
# Clean Each Trash Wheel Dataset and combine Datasets
```{r}
mr_trash   <- clean_trash("Mr. Trash Wheel", "Mr. Trash Wheel")
prof_trash <- clean_trash("Professor Trash Wheel", "Professor Trash Wheel")
gwynnda    <- clean_trash("Gwynns Falls Trash Wheel", "Gwynnda")
all_trash <- bind_rows(mr_trash, prof_trash, gwynnda)
```
# Compute Results
```{r}
n_obs <- nrow(all_trash)
prof_total_weight <- sum(prof_trash$weight_tons, na.rm = TRUE)
gwynnda_cig_2022_06 <- gwynnda |>
  filter(year == 2022, month == "June") |>
  summarise(total_cig = sum(cigarette_butts, na.rm = TRUE)) |>
  pull(total_cig)
n_obs
prof_total_weight
gwynnda_cig_2022_06
```

The combined Trash Wheel dataset contains `r n_obs` observations.  
Professor Trash Wheel collected a total of `r round(prof_total_weight, 2)` tons of trash.  
In June 2022, Gwynnda collected `r format(gwynnda_cig_2022_06, big.mark = ",")` cigarette butts.
Key variables include dumpster ID, collection date, total weight (in tons), counts of plastic bottles, polystyrene, cigarette butts, and sports balls.  

# Problem 3
library(readr)
library(dplyr)
library(tidyr)
library(janitor)

# Load and clean Zip Codes and ZORI data
```{r}
zips <- read_csv("data/Zip Codes.csv") %>% 
  clean_names()

zori <- read_csv("data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv") %>% 
  clean_names()
```
# Tidy ZORI data
```{r}
zori_tidy <- zori %>% 
  pivot_longer(
    cols = starts_with("x"),       
    names_to = "month",
    values_to = "rent_index"
  ) %>% 
  mutate(
    month = gsub("^x", "", month),                 
    month = as.Date(month, format = "%Y_%m_%d")    
  )
```
# Deduplicate ZIPs
```{r}
zips_unique <- zips %>% distinct(zip_code, .keep_all = TRUE)
```






